# The Major Weather Event Contributors of Health and Economic Impact within the United States

### This project looks at different weather events to determine the top 5 impactful events in areas of deaths, injuries, crop damage (in millions of dollars) and property damage (in millions of dollars). Key findings:
- Tornadoes are by far and away the largest contributor to deaths and fatalities
- Property damage is much greater in economic value than crop damage. 
- Droughts are the largest problem for crop damage whereas flooding creates the greatest property damage


## Loading Libraries
```{r loading libraries}
library(data.table)
library(dplyr)
library(ggplot2)
library(patchwork)
library(egg)
```


## DATA PROCESSING

In this section I load the data from the website and convert the property damage and crop damage fields to 
dollar amounts in millions for analyses. Currently the numeric columns for these are stored with different magnitudes
(single dollars[], thousands [K], millions [M], billions[B])

The data is also grouped by event type to calculate the total impact per area of interest (deaths, injuries,
crop damage and property damage). Finally, the top 10 are kept for report out.

```{r reading data}

Starting <- fread("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2")

Starting2 <- Starting %>% 
  mutate(propdmgM = case_when(PROPDMGEXP == 'K' ~ PROPDMG/1000,
                              PROPDMGEXP == 'M' ~ PROPDMG,
                              PROPDMGEXP == 'B' ~ PROPDMG*1000,
                              PROPDMGEXP == '' ~ PROPDMG/1000000
                              ),
         cropdmgM = case_when(CROPDMGEXP == 'K' ~ CROPDMG/1000,
                              CROPDMGEXP == 'M' ~ CROPDMG,
                              CROPDMGEXP == 'B' ~ CROPDMG*1000,
                              CROPDMGEXP == '' ~ CROPDMG/1000000
                              )) %>% 
  mutate(DeathInjuries = FATALITIES + INJURIES,
         CropPropDamageM = propdmgM + cropdmgM)

ByEventType <- Starting2 %>% 
  group_by(EVTYPE, dig.lab = 8) %>% 
  summarise(FATALITIES = sum(FATALITIES, na.rm = TRUE),
            INJURIES = sum(INJURIES, na.rm = TRUE),
            CROPDAMAGE = sum(cropdmgM, na.rm = TRUE),
            PROPDAMAGE = sum(propdmgM, na.rm = TRUE),
            CropPropDamageM = round(sum(CropPropDamageM, na.rm = TRUE), digits = 3),
            FatalitiesInjuries = sum(DeathInjuries)) %>% 
  ungroup()

FatalitiesTop <- ByEventType %>% 
  arrange(desc(FATALITIES)) %>% 
  top_n(5, FATALITIES)

InjuriesTop <- ByEventType %>% 
  arrange(desc(INJURIES)) %>% 
  top_n(5, INJURIES)

FatalitiesInjuriesTop <- ByEventType %>% 
  arrange(desc(FatalitiesInjuries)) %>% 
  top_n(5, FatalitiesInjuries)

PropDamageTop <- ByEventType %>% 
  arrange(desc(PROPDAMAGE)) %>% 
  top_n(5, PROPDAMAGE)

CropDamageTop <- ByEventType %>% 
  arrange(desc(CROPDAMAGE)) %>% 
  top_n(5, CROPDAMAGE)

CropPropDamageTop <- ByEventType %>% 
  arrange(desc(CropPropDamageM)) %>% 
  top_n(5, CropPropDamageM)



```

## RESULTS

I graph the results with a horizontal bar chart and only kept the top 5 weather events for both
health and economic impact.  Below you see the graphs for fatalities and injuries.  The main point is 
that tornadoes are the main culprit for deaths and injuries amongst the weather events.

```{r results1}

#png(filename="Injuries and Fatalities.png", width=1200, height=600)

p1 <- ggplot(FatalitiesTop, aes(x=reorder(EVTYPE, FATALITIES), y=FATALITIES)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
ggtitle('Top 10 for Fatalities') +
  ylab('Fatalities') +
  xlab('') +
  scale_y_continuous( limits=c(0, 100000))

p2 <- ggplot(InjuriesTop, aes(x=reorder(EVTYPE, INJURIES), y=INJURIES)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
ggtitle('Top 10 for Injuries') +
  ylab('Injuries') +
  xlab('Weather Event Type') +
  scale_y_continuous( limits=c(0, 100000))

p3 <- ggplot(FatalitiesInjuriesTop, aes(x=reorder(EVTYPE, FatalitiesInjuries), y=FatalitiesInjuries)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
ggtitle('Top 10 for Fatalities/Injuries Combined') +
  ylab('Fatalities/Injuries') +
  xlab('') +
  scale_y_continuous( limits=c(0, 100000)) +
  labs(caption = "Figure 1: Major Health Weather Events") +
  theme(plot.caption = element_text(hjust = 0))

p1+p2+p3 + plot_layout(nrow = 3) 

```


Below you see the economic impact of weather events in the form of crop and property damage.  Drought is
the major contributor to costs associated with crop damage.  However, the overall amount of damage from
crops is far less than the costs associated with property damage.  For property damage, flooding is the 
greatest contributor to economic impact.

```{r results2}

p4 <- ggplot(CropDamageTop, aes(x=reorder(EVTYPE, CROPDAMAGE), y=CROPDAMAGE)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
 ggtitle('Top 10 for Crop Damage') +
  ylab('') +
  xlab('') +
  scale_y_continuous( limits=c(0, 175000))

p5 <- ggplot(PropDamageTop, aes(x=reorder(EVTYPE, PROPDAMAGE), y=PROPDAMAGE)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
 ggtitle('Top 10 for Property Damage') +
  ylab('') +
  xlab('Weather Event Type') +
  scale_y_continuous( limits=c(0, 175000))

p6 <- ggplot(CropPropDamageTop, aes(x=reorder(EVTYPE, CropPropDamageM), y=CropPropDamageM)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
 ggtitle('Top 10 Crop and Property Damage Combined') +
  ylab('Damage ($Millions)') +
  xlab('') +
  scale_y_continuous( limits=c(0, 175000)) +
  labs(caption = "Figure 2: Major Economic Weather Events") +
  theme(plot.caption = element_text(hjust = 0))

p4+p5+p6 + plot_layout(nrow = 3) 





```


